An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-03-16 paper.tex
Merged Ilya's changes to intro.
First few paragraphs of IV.

A Markov chain has the property that the probability distribution of
the next state can depend only on the current state, not on the past
history: 
%
\be 
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to
\vtheta^{(j+1)}) p(\vtheta^{(j)}), \ee
%
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$
depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$.  An additional
requirement for an MCMC arises from the fact that the desired
distribution is the equilibrium distribution.  In other words, if we
assume that state $(j)$ of the chain is sampled from the desired PDF,
$p(\vtheta^{(j)})=p(\vtheta|d,M)$ , then the next state $(j+1)$ must
be sampled from the PDF as well, so that
$p(\vtheta^{(j+1)})=p(\vtheta|d,M)$; this condition is known as
``detailed balance.''

@@

This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can
depend on the parameters of the current state $\vtheta^{(j)}$, but not
on the past history.  It must also allow any state within the allowed
prior volume to be reachable by the MCMC.  Any jump proposal that
satisfies these properties is suitable for an MCMC.

The jump proposal is the most important choice in the MCMC, as it
determines the sampling efficiency of the algorithm, i.e., the
necessary length of the chain before it converges to the posterior
PDF.  Creating an efficient jump proposal distribution requires an
understanding of the structure of the parameter space which may not be
available until the PDFs are found, creating a Catch-22; one
possibility for resolving this infinite loop is described in Section
\ref{sec:examples}.

@@

The parameter space of the super-model in an RJMCMC consists of a
discrete parameter that identifies the model, $M_i$, and a set of
continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
each sample consists of a model identifier and a location within the
parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
MCMC in the ``super-model'' parameter space just like a regular MCMC;
in general, we can both propose jumps to different parameters within a
model (intramodel jumps) and jumps to a different model with different
parameters (intermodel jumps).  The resulting chain samples from the
posterior $p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the
model as a parameter, with other parameters ignored, is obtained by
marginalizing over the remaining parameters.  If we assume uniform
priors on the models, the evidence for a given model is proportional
to the number of counts 
%
\be
p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
\ee
%
where $N_i$ is the number of RJMCMC samples listing the $i$'th model
and $N$ is the total chain length.  Thus, the probability of a
particular model relative to other models under consideration is given
by the fraction of RJMCMC samples lying in the parameter space of that
model.

@@

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
We choose these neighborhoods to be non-overlapping and fill parameter
space.  To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  In other words, the size
of a neighborhood is inversely proportional to the local point
density.  The proposed jumps are drawn from a piecewise-constant
(i.e.\ constant on each neighborhood) interpolation of the PDF.  There
are various techniques that could be used to construct the set of
neighborhoods associated with each point.

Ref.~\cite{Littenberg2009} decomposed parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each point was associated with the brick that
contained it.  Unfortunately, this process does not guarantee that
every brick contains at least one point, which is required for the
jump proposal to cover the entire model parameter space.  The solution
to this problem was to add a uniform distribution to the proposal PDF,
in the form of one ``fictitious'' point per brick.  The extra uniform
distribution from the fictituous points represents a significant cost,
since inter-model jumps proposed from the uniform distribution are
unlikely to be accepted.  Additionally, the bricks in this algorithm
do not adapt to the local structure of the PDF.  One must either use
small bricks to capture the local structure of the PDF, placing many
bricks in regions without MCMC samples, or use large bricks, missing
the local strucutre of the PDF in exchange for fewer empty bricks.

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.  A
larger difficulty with the Voronoi approach is that the computational
cost for computing the Voronoi regions increases rapidly with
dimension.  % Apparently the exact computational cost is not known in
            % arbitrary dimension.

We propose to use a decomposition of parameter space into
neighborhoods based on a datastructure called a kD-tree (see, e.g.\
\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
efficient in high dimension than the Voronoi decomposition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{kdtree.png}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (which is also the number
    of boxes).}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate as a
    function of boxing number for the RJMCMC described in the text.}
\end{figure}

>>>>

A Markov chain has the property that the probability distribution of
the next state can depend only on the current state, not on the past
history: 
%
\be 
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to
\vtheta^{(j+1)}) p(\vtheta^{(j)}), \ee
%
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$
depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$.  An additional
requirement for an MCMC arises from the fact that the desired
distribution is the equilibrium distribution.  In other words, if we
assume that state $(j)$ of the chain is sampled from the desired PDF,
$p(\vtheta^{(j)})=p(\vtheta^{(j)}|d,M)$ , then the next state $(j+1)$ must
be sampled from the PDF as well, so that
$p(\vtheta^{(j+1)})=p(\vtheta^{(j+1)}|d,M)$; this condition is known as
``detailed balance.''

@@

This jump proposal distribution $Q(\vtheta^{(j)} \to \vtheta^p)$ can
depend on the parameters of the current state $\vtheta^{(j)}$, but not
on the past history.  It must also allow any state within the prior
volume to be reachable (eventually) by the MCMC.  Any jump proposal
that satisfies these properties is suitable for an MCMC.

The jump proposal is the most important choice in the MCMC, as it
determines the sampling efficiency of the algorithm, i.e., the length
of the chain before it converges to the posterior PDF.  Creating an
efficient jump proposal distribution requires an understanding of the
structure of the parameter space which may not be available until the
PDFs are found, creating a Catch-22; one possibility for resolving
this infinite loop is described in Section \ref{sec:examples}.

@@

The parameter space of the super-model in an RJMCMC consists of a
discrete parameter that identifies the model, $M_i$, and a set of
continuous parameters appropriate for that model, $\vtheta_i$.  Thus,
each sample consists of a model identifier and a location within the
parameter space of that model, $\{M_i, \vtheta_i\}$.  We perform the
MCMC in the ``super-model'' parameter space just like a regular MCMC;
we propose jumps to different parameters within a model (intramodel
jumps) and jumps to a different model with different parameters
(intermodel jumps).  The resulting chain samples from the posterior
$p(M_i,\vtheta_i|d)$.  As in a usual MCMC, the PDF on the model as a
parameter, with other parameters ignored, is obtained by marginalizing
over the remaining parameters.  The posterior probability of a model
is proportional to the number of counts
%
\be
%
p(M_i|d) = \int d\vtheta_i \frac{p(d|M_i, \vtheta_i) p(M_i, \vtheta_i)}{p(d)}
\approx \frac{N_i}{N},
%
\ee
%
where $N_i$ is the number of RJMCMC samples listing the $i$'th model
and $N$ is the total chain length.  Thus, the probability of a
particular model relative to other models under consideration is given
by the fraction of RJMCMC samples lying in the parameter space of that
model.

@@

The problem of drawing a proposed jump from an interpolation of
single-model MCMC points can be thought of as the problem of assigning
a local ``neighborhood'' to each point in the chain of MCMC samples.
We choose these neighborhoods to be non-overlapping and fill parameter
space.  To draw a proposed jump, we select a point uniformly from the
MCMC samples, find its associated neighborhood, and then draw the
proposed jump uniformly from the neighborhood.  Since the MCMC points
are distributed according to the posterior PDF for the single model,
this procedure produces proposed jumps that are approximately
distributed according to the posterior PDF.  The size of a
neighborhood is inversely proportional to the local point density.
The proposed jumps are drawn from a piecewise-constant (constant on
each neighborhood) interpolation of the PDF.  There are various
techniques that could be used to construct the set of neighborhoods
associated with each point.

Ref.~\cite{Littenberg2009} decomposed parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each point was associated with the brick that
contained it.  Unfortunately, this process does not guarantee that
every brick contains at least one point, which is required for the
jump proposal to cover the entire model parameter space.  The solution
to this problem was to add a uniform distribution to the proposal PDF,
in the form of one ``fictitious'' point per brick.  The extra uniform
distribution from the fictituous points can represent a significant
cost, since inter-model jumps proposed from the uniform distribution
are unlikely to be accepted.  Additionally, the bricks in this
algorithm do not adapt to the local structure of the PDF.  One must
either use small bricks to capture the local structure of the PDF,
placing many bricks in regions without MCMC samples, or use large
bricks, missing the local strucutre of the PDF in exchange for fewer
empty bricks.

An alternate technique for producing adaptive neighborhoods would be
to use the Voronoi regions associated with each MCMC sample.  The
Voronoi region associated with a sample contains all the parameter
space points that are closer to that sample than any other sample.
The Voronoi region decomposition into neighborhoods is, in a sense,
maximally adaptive, in contrast to the approach of
Ref.~\cite{Littenberg2009}, which is minimally adaptive.
Unfortunately, defining the Voronoi regions requires a metric on
parameter space, which may be difficult or impossible to define.
Also, the computational cost for computing the Voronoi regions
increases rapidly with
dimension.  % Apparently the exact computational cost is not known in
            % arbitrary dimension.

Here we propose to use a decomposition of parameter space into
neighborhoods based on a datastructure called a kD-tree (see, e.g.\
\cite{Berg2008} or \cite{Gaede1998}).  The decomposition is more
adaptive than the cells of Ref.~\cite{Littenberg2009}, and more
efficient in high dimension than the Voronoi decomposition.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{kdtree.png}
  \end{center}
  \caption{\label{fig:kD-tree} The neighborhoods from a kD-tree
    constructed around a set of points that are normally distributed
    about the origin in two dimensions.  As the points become denser
    around the origin, the typical neighborhood gets smaller.  The
    interpolated PDF within a box of volume $V_i$ is $1/(N V_i)$,
    where $N$ is the total number of points (which is also the number
    of boxes).}
\end{figure}

@@

In this section, we demonstrate the efficiency of the kD-interpolated
jump proposal on a toy model comparison problem.  We draw $N = 100$
simulated data points from a $N(0,1)$ Gaussian distribution, and then
ask whether these data are better described by a model where they are
Gaussian distributed with unkonwn mean, $\mu$, and standard deviation,
$\sigma$
%
\be
p(x) = \frac{1}{\sqrt{2\pi} \sigma} \exp\left( - \frac{(x-\mu)^2}{2
    \sigma^2} \right),
\ee
%
or by a model where they are Cauchy distributed with mode $\alpha$,
and width $\beta$
%
\be
p(x) = \frac{1}{\pi \beta \left( 1 + \left(\frac{x - \alpha}{\beta}\right)^2\right)}.
\ee
%
We take priors on $\mu$ and $\alpha$ to be uniform in $[-1,1]$, and
priors in $\sigma$ and $\beta$ to be uniform in $[0.5, 1.5]$.  With a
data set of 100 points, the relative uncertainty in determining the
parameters of the underlying distribution is approximately 10\%, so we
expect the posterior probabilities in the $(\mu,\sigma)$ and
$(\alpha,\beta)$ spaces to occupy roughly 1\% of the prior volume.

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate as a
    function of boxing number for the RJMCMC described in the text.}
\end{figure}
